---
title: "Comparing Social Mobility across Countries with the UNIDIFF Model"
output: github_document
params:
  data: 
    label: Path to data file
    value: NULL
    input: file
  artefacts: 
    label: Save artefacts (simulated data and intermediate objects)?
    value: FALSE
    input: file
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = "")
saved_results <- is.null(params$data)
if (saved_results){
    for (nm in c("dev", "df", "periodContr", "classContr",
                 "modelC", "modelD", "modelEa", "modelJ")){
        assign(nm, readRDS(file.path("UNIDIFF_artefacts", "saved_results",
                                     paste0(nm, ".rds"))))
    }
}
```

## Loading required packages

```{r packages}
library(gnm)
library(vcdExtra)
```

## Pre-processing the Data

The survey data is read into R, selecting the variables of interest: the five-year period, the eight-category social class variable for father and respondent, and the weighted frequencies. The period and social class variables are converted to factors.


```{r read-data, eval = !saved_results}
allMen <- read.table(params$data, header = FALSE)[, c(4, 10, 12, 14)]
colnames(allMen) <- c("period", "origin8", "dest8", "wFreq")
allMen[,1:3] <- lapply(allMen[, 1:3], factor)
```

Following the analysis by Ganzeboom and Luijkx (in Social Mobility in Europe, ed. R. Breen, 2004), we collapse the data by the factors `period`, `origin8`, and `dest8`.

```{r dataI, eval = !saved_results}
dataI <- as.data.frame(xtabs(wFreq ~ period + origin8 + dest8, data = allMen))
```

Since the response is a weighted frequency, the resulting counts are non-integer. This generates multiple warnings during the model fitting, which are suppressed in the output below.
## Modelling
We first fit the conditional independence model as a baseline (the model names refer to the names used in Ganzeboom and Luijkx, 2004):

```{r indep, eval = !saved_results}
modelA <- suppressWarnings(gnm(Freq ~ -1 + period:origin8 + period:dest8,
                               family = poisson, data = dataI, verbose = FALSE))
```

Then we update this model to the unstructured UNIDIFF model, using `Exp` to ensure the period multipliers are non-negative and the `ofInterest` argument to mark these parameters as the only ones that should be included in model summaries.

```{r unstructured, eval = !saved_results}
modelC <- update(modelA, . ~ . + Mult(Exp(period), origin8:dest8),
                 ofInterest = "[.]period")
dev <- c("C" = deviance(modelC))
df <- c("C" = df.residual(modelC))
```
```{r modelC-stats}
dev["C"]
df["C"]
```

The `getContrasts()` function is used to obtain simple contrasts of the log-multipliers along with their quasi-variances:

```{r calc-periodContr, eval = !saved_results}
periodContr <- getContrasts(modelC, ofInterest(modelC))
```
```{r summary-periodContr}
summary(periodContr)
```

We then consider two models suggested by Ganzeboom and Luijkx, with either a linear or quadratic dependence of the multipliers on period number.

```{r multiplier-dependence, eval = !saved_results}
modelD <- update(modelA, . ~ . + Mult(Const(1) + as.numeric(period),
                                      origin8:dest8),
                 ofInterest = "[.]as")
modelE <- update(modelA, . ~ . +
                 Mult(Const(1) + as.numeric(period)+ I(as.numeric(period)^2),
                      origin8:dest8),
                 ofInterest = "[.]as")
dev <- c(dev, "D" = deviance(modelD), "E" = deviance(modelE))
df <- c(df, "D" = df.residual(modelD), "E" = df.residual(modelE))
```
```{r modelDE-stats}
dev[c("D", "E")]
df[c("D", "E")]
```

Finally we propose a model in which the log-multipliers depend on period number -- note a `Const` term is unnecessary here as this would simply scale the multipliers by a constant.

```{r log-multiplier-dependence, eval = !saved_results}
modelEa <- update(modelA, . ~ . + Mult(Exp(as.numeric(period)), origin8:dest8),
                  ofInterest = "[.]as")
dev <- c(dev, "Ea" = deviance(modelEa))
df <- c(df, "Ea" = df.residual(modelEa))
```
```{r modelEa-stats}
dev["Ea"]
df["Ea"]
```

Having considered a selection of models, we impose some identifiability constraints, for presentational purposes. For the UNIDIFF model, we constrain the log-multiplier for the first period to 0:

```{r constrain-modelC, eval = !saved_results}
modelC <- update(modelC, constrain = "[.]period1", constrainTo = 0)
```

For the structured UNIDIFF models, we fix the origin-destination interaction as in `modelC` - note that this constraint slightly impairs model fit, but enables the multipliers to be directly compared as in the plot below.

```{r constrain-D-and-Ea, eval = !saved_results}
keep <- pickCoef(modelC, "[.]origin")
modelD <- update(modelD, constrain = "[.]origin",
                 constrainTo = coef(modelC)[keep],
                 ofInterest = "[.]as")
modelEa <- update(modelEa, constrain = "[.]origin",
                  constrainTo = coef(modelC)[keep],
                  ofInterest = "[.]as")
```

A plot of the period contrasts from the UNIDIFF model and a second plot of the corresponding scaling factors are created as follows. The points are joined by lines in the second plot to highlight the trends in the estimates.

```{r periodContr, tidy = FALSE, dev = "png", fig.width = 900/72, fig.height = 500/72, out.width = "100%", fig.alt = "contrasts of period multipliers"}
par(mfrow = c(1,2), mar = c(4.5, 4.1, 3.1, 1.1), cex = 0.9, mex = 0.8,
    oma = c(0, 0, 0, 0))
plot(periodContr, xaxt = "n", xlab = "Period",
     ylab = "Log-multipliers of origin-destination association",
     main = "")
axis(1, lab = c("1970-4", "1975-9", "1980-4", "1985-9", "1990-4", "1995-9"),
     at = 1:6)
mtext("a)", side = 2, line = 2, at = 0.19, las = 2)
plot(1:6, exp(parameters(modelC)[ofInterest(modelC)]), ylim = c(0.6, 1.0),
     xlab = "Period number",
     ylab = "Multipliers of origin-destination association")
lines(1:6, 1 + coef(modelD)[ofInterest(modelD)]*1:6, lty = 2)
lines(1:6, exp(coef(modelEa)[ofInterest(modelEa)]*1:6), lty = 1)
mtext("b)", side = 2, line = 2, at = 1.03, las = 2)
```

We finally consider the Goodman-Hauser model selected by Ganzeboom and Luijkx:

```{r GHmodel, eval = !saved_results}
modelJ <- update(modelA, . ~ . + Diag(origin8, dest8) +
                 Mult(Const(1) + as.numeric(period), MultHomog(origin8, dest8)),
                 etastart = modelA$predictors)
dev <- c(dev, "J" = deviance(modelJ))
df <- c(df, "J" = df.residual(modelJ))
```
```{r modelJ-stats}
dev["J"]
df["J"]
```

This time we focus on the homogenous multiplicative effects of origin/destination class, using `getContrasts` to obtain the simple contrasts of these parameters for summary and display:

```{r class-contrasts, eval = !saved_results}
classContr <- getContrasts(modelJ, pickCoef(modelJ, "[.]origin"))
```
```{r summary-classContr}
summary(classContr)

```
```{r classContr, dev = "png", fig.width = 600/72, fig.height = 500/72, out.width = "100%", fig.alt = "contrasts of class multipliers"}
plot(classContr, xaxt = "n", xlab = "Social class",
     ylab = "Multiplicative effects of origin/destination class",
     main = "")
axis(1, lab = c("I", "II", "III", "IVab", "IVc", "V + VI", "VIIa", "VIIb"),
     at = 1:8)
```

We evaluate the fit of this model by plotting the aggregate residuals:

```{r GHmosaic, dev = "png", fig.width = 570/72, fig.height = 475/72, out.width = "100%", fig.alt = "aggregate residuals from Goodman-Hauser model"}
mosaic(modelJ, ~origin8 + dest8,
       set_varnames =
       list(origin8 = "Origin class", dest8 = "Destination class"),
       set_labels =
       list(origin8 = c("I", "II", "III", "IVab", "IVc", "V + VI", "VIIa", "VIIb"),
       dest8 = c("I", "II", "III", "IVab", "IVc", "V + VI", "VIIa", "VIIb")),
       rot_labels = c(0, 0), offset_varnames = c(0.3, 0, 0, 1.3),
       offset_labels = c(0, 0, 0, 0.3), margins = c(1, 1, 0, 4))
```

## References
Ganzeboom, Harry B. G., and Ruud Luijkx. 2004. “Recent trends in intergenerational occupational class reproduction in the Netherlands 1970-99.” In *Social Mobility in Europe*, edited by Richard Breen, 345–81. Oxford University Press.

```{r simulated, eval = params$artefacts, include = FALSE}
# simulate counts based on modelEa 
set.seed(1)
sim_freq <- simulate(modelEa, nsim = 1)[[1]]
# create dummy data.frame
dat <- matrix(NA, nrow = nrow(dataI), ncol = 14)
dat[,c(4, 10, 12)] <- vapply(dataI[, 1:3], unclass, numeric(nrow(dataI)))
dat[,14] <- sim_freq
# write files
output_dir <- "UNIDIFF_artefacts/simulated_data"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
write.table(dat, file.path(output_dir, "social_class.txt"), row.names = FALSE,
            col.names = FALSE)

```
```{r save-results, eval = params$artefacts, include = FALSE}
results_dir <- "UNIDIFF_artefacts/saved_results"
## keep coef info only for models C, D, Ea
coef_info <- function(mod){
    res <- mod[c("coefficients", "ofInterest", "constrain", "constrainTo")]
    class(res) <- class(mod)
    res
}
## keep info on modelJ to reproduce mosaic plot
mosaic_info <- function(mod){
    res <- mod[c("coefficients", "ofInterest", "constrain", "constrainTo",
                 "call", "df.residual", "na.action", "family")]
    ## aggregate over period, round as in Table 14.A2 of Ganz04
    res$data <- as.data.frame(xtabs(Freq ~ origin8 + dest8, data = dataI))
    res$data$Freq <- round(res$data$Freq, 1)
    ## fitted values can be put to anything here
    n <- length(res$data$Freq)
    res$fitted.values <- rep.int(100, n)
    ## fix residuals to correct values and add in dummy values
    ## (so it seems there is more than one value to average)
    zero <- res$data
    zero$Freq <- 0
    res$data <- rbind(res$data, zero)
    res$y <- res$data$Freq
    mean_resid <- meanResiduals(mod, by = ~origin8:dest8)
    res$residuals <- c(c(mean_resid)/sqrt(res$fitted.values), 
                       numeric(nrow(zero)))
    res$predictors <- c(log(res$fitted.values), rep(-Inf, n))
    res$fitted.values <- c(res$fitted.values, rep(1, n))
    res$prior.weights <- rep.int(1, 2*n)
    res$x <- matrix(1, nrow = 2*n) # dummy as rank has no impact
    class(res) <- class(mod)
    res
}

if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
for (nm in c("dev", "df", "periodContr", "classContr")){
    saveRDS(get(nm), file.path(results_dir, paste0(nm, ".rds")))
}
for (mod in c("modelC", "modelD", "modelEa")){
    saveRDS(coef_info(get(mod)), file.path(results_dir, paste0(mod, ".rds")))
}
saveRDS(mosaic_info(modelJ), file.path(results_dir, "modelJ.rds"))
```

